---
title: "Optim Example"
output: html_document
date: "2024-05-29"
---

```{r}
library(ggplot2)
library(kableExtra)
```


Here we will walk through an example using `optim` by using it to minimize the MSE (mean squared error) for a linear function, which in all honesty is over fit there are better functions there much better built in function that do this (aka `lm` which is used to fit linear models) but it is probably the simplest demo that comes to mind. 



First we create the data set we will use for this example: 

```{r}
# Set a random seed for reproducibility
set.seed(123)

# Create random data
x <- rnorm(500)
y <- rnorm(500) + 0.7 * x

# Combine x and y into a data frame
data <- data.frame(x, y)
```

```{r}
ggplot(data = data) + 
  geom_point(aes(y, y))
```

```{r}
# Manually create a function for residual sum of squares
# Args 
#   data: data.frame containing a x and y column 
#   par: numeric vector of two parameters 
# Rerun: numeric vector length 1 the MSE between predicted y and the observed y   
objective_function <- function(data, par) {
  
  predicted_y <- par[1] + par[2] * data$x
  SE <- (predicted_y - data$y)^2
  MSE <- mean(SE)
  return(MSE)
  
}
```


Use the `optim` function to estimate values of our two parameters. 
```{r}
# Applying optim
optim_output <- optim(par = c(0, 1),
                      fn = objective_function,
                      data = data)

```


Question for Peter what happens when you play aroud with the values of par? If you set par = (10, 10)? Try palying around with different values


```{r}
optim_output
```


Now estimate the parameters using `lm`

```{r}
fit <- lm(data = data, formula = y ~ x)
```

```{r}
fit
```


Questions for Peter 

* What is returned by optim? What kind of object is it? What do the different elements refer to? What about the lm function? 
* How different were the parameter estimates? Can compare in a nice table (use`kable`)?
* How do the predicted results using the different par estimates compare with one another? 
* Evaluate parameter fit performance (compare predicted values of y with the original values using MSE)



