---
title: "Hector [CO2] vs. Observations"
output: html_document
author: "Peter Scully"
date: "2024-05-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting the data

First, let's import the relevant libraries and setup our directory location
```{r import_and_loc}
library(hector)
library(ggplot2)
library(dplyr)

INPUT_DIR <- file.path(here::here(), '01')
```

Now, let's read the historical data into a dataframe

```{r hist_table}
hist_data <- read.table(file.path(INPUT_DIR, 'co2_annmean_mlo.txt'), 
                        skip = 43, 
                        col.names = c("year", "mean", "unc"))
hist_data
```

We also need to get the default Hector data:

```{r hector_run}
# Setting the desired date range and variables
yrs <- 1959:2023
hector_vars <- CONCENTRATIONS_CO2()

# Running the default ssp245 run
ini_file <- system.file("input/hector_ssp245.ini", package = "hector")
core <- newcore(ini_file)
run(core)
hector_data <- fetchvars(core, yrs, vars = hector_vars)

hector_data

```

## Making the comparison plot

Now that we have both the historical and model data, we need to combine the
data frames so we can make a comparison plot:
```{r combining data frames}

# First, let's delete the unnecessary columns
hist_data <- subset(hist_data, select = c('year', 'mean'))
hector_data <- subset(hector_data, select = c('year', 'value'))

# Now, let's make the column names match so we can combine the data frames
colnames(hist_data) <- c('year', 'CO2_concentration')
colnames(hector_data) <- c('year', 'CO2_concentration')

# Finally, let's add a column to distinguish which data comes from which source
hist_data$data_source <- "Historical Record"
hector_data$data_source <- "Hector"

# Combining the data frames
combined_data <- rbind(hist_data, hector_data)

```

With this new data frame, we can now make the actual comparison plot

```{r plotting}
plot <- ggplot(data = combined_data, 
               aes(x = year, y = CO2_concentration, color = data_source)) +
  geom_line() + 
  ggtitle("Comparing Results")
plot
```


## Calculating mean squared error

Now we just need to calculate mean squared error. We'll start by just creating
a generic function to calculate MSE:
```{r mse_fxn}
mse <- function(x, y) {
  squared_error = (x - y) ^ 2
  sum(squared_error) / length(squared_error)
}
```

With this function, we can calculate the MSE between real observations and the
results from Hector:
```{r final_mse}
final_mse = mse(hector_data$CO2_concentration, hist_data$CO2_concentration)
```

Thus, the MSE between the model results and historical observations is 
`r final_mse`.
