---
title: "Running Hector with Constraints"
output: html_document
date: "2024-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
To begin, we'll need to import some libraries and set some initial variables:
```{r imports_consts, warning=FALSE, message=FALSE}
library(hector)
library(ggplot2)
library(dplyr)

hector_vars <- c(GMST(), CONCENTRATIONS_CO2())
yrs <- 1745:2300
```

## Default Run

First, let's run the default emission-driven ssp245 run:

```{r default}
ini_file <- system.file("input/hector_ssp245.ini", package = "hector")
core <- newcore(ini_file)
run(core)
initial_data <- fetchvars(core, yrs, vars = hector_vars)
```

## Setting Constraints via the R Interface

First, let's read in the CO2 constraint data
```{r reading_co2}
constr_df <- read.csv(system.file("input/tables/ssp245_emiss-constraints_rf.csv",
                                  package = "hector"),
                      skip = 5)
CO2_constr <- constr_df$CO2_constr
CO2_constr
```

Now, let's run Hector constrained by this CO2 data
```{r interface}
# Adding in the constraint
setvar(core, 
       dates = yrs, 
       var = CO2_CONSTRAIN(), 
       values = CO2_constr,
       unit = getunits(CO2_CONSTRAIN()))
reset(core)

# Running Hector with the constraint
run(core)
constr_data_interface <- fetchvars(core, yrs, vars = hector_vars)

constr_data_interface
```


## Setting Constraints via the .ini file
After editing the .ini file to run Hector with CO2 constraints, we can now run
Hector using the modified ini file
```{r ini}
# Creating the new core
constr_ini_file <- file.path(here::here(), 
                             '01', 
                             'hector_inputs', 
                             'hector_inputs',
                             'hector_ssp245.ini')
constr_core <- newcore(constr_ini_file)

# Running the new core
run(constr_core)
constr_data_ini <- fetchvars(constr_core, yrs, vars = hector_vars)
constr_core <- shutdown(constr_core)

constr_data_ini
```


## Comparing Results

Now that we've completed all 3 runs, we can compare their results. To do this,
we first need to combine their data frames:
```{r combine_dfs}
# Updating scenario names in data frames
initial_data$scenario <- "emission driven run"
constr_data_interface$scenario <- "constrained via R interface"
constr_data_ini$scenario <- "constrained via .ini file"

# Combining data frames
combined_data <- rbind(initial_data, constr_data_interface, constr_data_ini)
```

Now, let's make our final plot:
```{r plotting}
ggplot() + 
  geom_line(data = combined_data, 
            aes(x = year, y = value, color = scenario, linetype =  scenario)) +
  facet_wrap(~ variable, scales = "free") +
  ggtitle("Constrained and Unconstrained Hector Runs")
  
```

As expected, the two methods of constraining the data produce the same results

## Bonus: Changing Beta and Q10

Let's see what happens when we change beta and Q10 (using the core constrained via the R interface):
```{r beta_q10, results='hide',fig.keep='all'}
# Changing beta
setvar(core, dates = NA, var = BETA(), values = 0.72, unit = getunits(BETA()))
reset(core)
run(core)
new_beta_data <- fetchvars(core, yrs, vars = hector_vars)
new_beta_data$scenario <- "constrained, beta = 0.72"

# Changing Q10
setvar(core, dates = NA, var = Q10_RH(), values = 4.0, unit = getunits(Q10_RH()))
reset(core)
run(core)
new_q10_data <- fetchvars(core, yrs, vars = hector_vars)
new_q10_data$scenario <- "constrained, Q10 = 4.0"

shutdown(core)

# Combining data frames
bonus_data <- rbind(constr_data_interface, new_beta_data, new_q10_data)

# Making plot
ggplot() + 
  geom_line(data = bonus_data, 
            aes(x = year, y = value, color = scenario, linetype =  scenario)) +
  facet_wrap(~ variable, scales = "free") +
  ggtitle("Constrained Hector Runs with Different Beta and Q10")

```

These plots indicate that changing beta and Q10 did not affect the results of the constrained runs. This makes sense because, during a constrained run, Hector overwrites CO2 data from the carbon cycle with the prescribed data, making any carbon cycle parameters irrelevant.
