---
title: "Hector v. Observed Temperatures"
author: "Peter Scully"
date: "2024-06-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Data

To start, let's import the relevant libraries:

```{r imports, message=FALSE, warning=FALSE}
library(hector)
library(ggplot2)
library(dplyr)
```

We now need to import the historical data into a data frame
```{r cars}
obs_data <- 
  read.csv(file.path(here::here(), "01", 
                     "HadCRUT.5.0.2.0.analysis.summary_series.global.annual.csv"))
colnames(obs_data) <- c("year", "temperature", "lower", "upper")

obs_data
```

We also need to run Hector:
```{r}
# Setting the desired date range and variables
yrs <- 1850:2024               # Date range of observational record
hector_vars <- GLOBAL_TAS()    # Observed temperatures are near-surface

# Running the default ssp245 run
ini_file <- system.file("input/hector_ssp245.ini", package = "hector")
core <- newcore(ini_file)
run(core)
hector_data <- fetchvars(core, yrs, vars = hector_vars)
shutdown(core)

hector_data
```

## Converting Data

Right now, the Hector data contains temperature anomalies relative to 1750 but
the observational data contains temperature anomalies relative to 1961-1990. To
resolve this, let's write some functions to convert Hector's temperatures.

We can start by creating a function to get the average temperature over a given interval

```{r}
#######################
# get_interval_avg - function to find the average value of a Hector variable over
#                    a given time interval
# args:
#   data - data frame outputted by fetchvars
#   var - string containing the Hector variable to average
#   lower - start of time interval to average across
#   upper - end of time interval to average across
#
# returns - average value of var over the provided interval
#######################
get_interval_avg <- function(data, var, lower, upper) {
  data %>%
    subset(variable == var) %>%
    subset(lower <= year & year <= upper) %>%
    .$value -> interval_vals
  return(mean(interval_vals))
}

hector_avg_T = get_interval_avg(hector_data, 
                                var = GLOBAL_TAS(), 
                                lower = 1961, 
                                upper = 1990)

hector_avg_T
```

Next, we need a function to use this average to adjust Hector's temperature
values. We'll add these values to a new data frame containing the Hector
data we'll use for our comparison

```{r}
#######################
# new_rel_val - function to adjust values of a Hector variable to be relative
#               to a new value
# args:
#   data - data frame outputted by fetchvars
#   var - string containing the Hector variable to adjust
#   benchmark - value to make variable relative to 
#
# returns - vector containing new values for var
#######################
new_rel_val <- function(data, var, benchmark) {
  data %>%
    subset(variable == var) %>%
    .$value - benchmark -> new_vals
  return(new_vals)
}

hector_adjusted <- data.frame(hector_data$year, 
                              new_rel_val(data = hector_data,
                                          var = GLOBAL_TAS(),
                                          benchmark = hector_avg_T))

# Making this data frame line up with obs_data
colnames(hector_adjusted) <- c("year", "temperature")
hector_adjusted$lower <- 0
hector_adjusted$upper <- 0

hector_adjusted

```


## Comparing Results
From here, we can now try to find the RMSE between the two temperature series.
To help with this, we can use another function:
```{r}
#######################
# rmse - function to find the root mean squared error between vectors of
#        predicted and observed values
#
# args:
#   x - vector of predicted values
#   y - vector of observed values (same length as x)
#
# returns - RMSE of x and y
#######################
rmse <- function(x, y) {
  SE = (x - y)^2
  MSE = mean(SE)
  return(sqrt(MSE))
}

rmse(hector_adjusted$temperature, obs_data$temperature)
```

Let's see how this value lines up with a plot comparing temperature predictions:
```{r}
# First we need to combine the data frames
hector_adjusted$source <- "Hector"
obs_data$source <- "Observations (HadCRUT.5.0.2.0)"
combined_data <- rbind(hector_adjusted, obs_data)

# Now we can make our plot:
ggplot(data = combined_data, aes(x = year, y = temperature, color = source)) +
  geom_ribbon(data = filter(combined_data, 
                            source == "Observations (HadCRUT.5.0.2.0)"),
              aes(ymin = lower, ymax = upper), 
              fill = 'aquamarine1', color = NA) +
  geom_line() +
  ggtitle("Comparing Results")
```
